---
title: "Expected Goals Prediction (xG)"
author: "Maha Salama"
date: "11/17/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The term xG in football stands for 'expected goals'. It is a statistical measurement of the quality of goalscoring chances and the likelihood of them being scored.

The goal of this task is to use shots and frames data to identify what could be the best predictors for goals, so that we have a new metric (xG) created, that calculates the probability of a shot being scored.

```{r echo = FALSE, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(corrplot)
library(sjmisc)
library(plotly)
library(reshape2)
```

### Data Loading:
First, the 2 dataframes used for this task are loaded, shots_df (containing shot data), and frames_df (containing freeze frame data that represents players other than the shooter with their locations and positions at the time of each shot).

```{r message=FALSE, warning=FALSE}
shots.df <- read.csv("./shots_df.csv")
frames.df <- read.csv("./shots_freeze_frames_df.csv")

```

### Data Exploration:
Data is explored and some summary statistics applied.
```{r message=FALSE, warning=FALSE}
summary(shots.df)
summary(frames.df)
```

### Data Cleaning:
NAs removed, categorical variables adjusted, 
Target variable added (is_goal): created from outcome.name variable.
```{r message=FALSE, warning=FALSE}
shots.df[is.na(shots.df)] <- FALSE

shots.df$is_goal <- 0
shots.df$is_goal[shots.df$shot.outcome.name == "Goal"] <- 1
```

### Feature Engineering:
The 2 dataframes are joined, and the needed features for the analysis are created.

##### Features to be used: 
1. Shot distance: Distance between the shot location and the central point of the goal line (120,40). 
2. Shot angle: calculated from shot location, dimensions of the field and length of the goal (7.32 meters)
3. opponents in space: how many opponents in between the player and the goal
4. defenders in space: how many of them are defenders
5. goalkeeper position
```{r message=FALSE, warning=FALSE}
cleaned.data <- shots.df %>%
  rename(x= location.x, y= location.y)%>%
  mutate(shot.distance = sqrt((120 - x)^2 + (40 - y)^2),
         shot.angle = atan(7.32*(120 - x)/((120 - x)^2+(40 - y)^2-(7.32/2)^2))* 180/pi) %>%
  modify_if(is.character, as.factor) 

# Join the 2 DFs at shot id, and extract features about opponents positions:
## (number of opponents in between the player and the goal, number of defenders, position of goalkeeper).
joined.df <- merge(cleaned.data, frames.df, by = "id") %>%
  rename(other.x= location.x, other.y= location.y, other.position = position.name.y)%>%
  mutate(other.distance = sqrt((120 - other.x)^2 + (40 - other.y)^2),
         other.in.space = ifelse(other.distance < shot.distance, TRUE, FALSE),
         opp.in.space = ifelse(other.in.space & !teammate, TRUE, FALSE),
         defenders.in.space = ifelse(opp.in.space & grepl("Defensive", other.position), TRUE, FALSE))

eng.df <- joined.df %>% 
  group_by(id,player.name.x,team.name,
           minute,second,possession,duration,under_pressure,
           play_pattern.name, shot.body_part.name,
           shot.technique.name, shot.type.name, match_id,
           shot.open_goal, shot.first_time, shot.redirect, shot.deflected, shot.saved_to_post, 
           shot.distance, shot.angle,shot.statsbomb_xg, is_goal) %>%
  summarize(sum.opp.in.space = sum(opp.in.space),
            sum.defenders.in.space = sum(defenders.in.space),
            goal_keeper_distance = other.distance[other.position == "Goalkeeper"])
```


### Data Exploration Again:
After feature engineering and joining the dataframes, data is explored again to discover relations and correlations between the variables are explored.
```{r echo = FALSE, message=FALSE, warning=FALSE}
data.cor <- eng.df[,-grep("match.id",colnames(eng.df))]
data.cor <- cor(data.cor[, unlist(lapply(data.cor, is.numeric))])
```
```{r cor, echo=FALSE, fig.cap="correlation matrix"}
corrplot(data.cor)
```

### Data Modeling:
Next, a model is created to make predictions of expected goals.
First, the data is split into train and test.
Then, logistic regression is applied to the data to generate probabilities of shots being scored.
The features used to build this model are:
shot.distance, shot.angle, sum.opp.in.space, sum.defenders.in.space, and goal_keeper_distance.
```{r message=FALSE, warning=FALSE}
# split the data into train & test
data <- eng.df
set.seed(101) 
# Selecting 80% as train data  
sample <- sample.int(n = nrow(data), size = floor(.8*nrow(data)), replace = F)
train <- data[sample, ]
test  <- data[-sample, ]

# Build the model: logistic regression
mod <- glm(is_goal ~ shot.distance  + shot.angle + sum.opp.in.space + sum.defenders.in.space + goal_keeper_distance, data=train, family=binomial)
summary(mod)

# getting the intercepts of the model
int <- coef(mod)
int_coef <- int[1]
dist_coef <- int[2]
ang_coef <- int[3]
opp_coef <- int[4]
def_coef <- int[5]
gk_coef <- int[6]

# giving the xG value to the shots
for (i in seq(1,nrow(data))){
  sum = int_coef + ang_coef*data[i,"shot.angle"] + (dist_coef*data[i,"shot.distance"])+ 
    (opp_coef*data[i,"sum.opp.in.space"])+ (def_coef*data[i,"sum.defenders.in.space"])+ 
    (gk_coef*data[i,"goal_keeper_distance"])
  data[i,"xG"] = exp(sum)/(1+exp(sum))
}
data
```

### Plotting and Visualization:
Next, some useful visualizations are created to plot the new metric (xG) against different variables and compare it to actual goals.

##### 1. xG versus shot distance
```{r echo = FALSE, message=FALSE, warning=FALSE}
# xG versus distance
xg_distance <- data %>% ggplot(aes(data$shot.distance,data$xG)) + 
  geom_point(color = "#20B2AA")+ ylim(c(0,1))+ geom_smooth(color = "orange")+
  labs(title= "xG vs Distance",x ="Distance", y = "xG")

```
```{r distance, echo=FALSE}
xg_distance
```

##### 2. xG vs statsBomb_xG
```{r echo = FALSE, message=FALSE, warning=FALSE}
# xG vs statsBomb_xG
xg_statsBomb <- data %>% ggplot(aes(data$xG,data$shot.statsbomb_xg)) + 
  geom_point(color = "#20B2AA")+ geom_abline(color = "orange") +ylim(c(0,1))+
  labs(title= "xG vs statsBomb_xG",x ="xG", y = "statsBomb xG")

```
```{r xG, echo=FALSE}
xg_statsBomb
```

##### 3. Shots distribution across xG: 
A plot visualizing shots spread versus xG to see where the majority of shots lie
```{r echo = FALSE, message=FALSE, warning=FALSE}
# shots distribution across xG
shots.distribution <- data %>% ggplot(aes(data$xG)) + 
  geom_histogram(color = "gray", fill = "#20B2AA")+
  labs(title= "Shots Distribution Across Expected Goals",x ="xG", y = "Shots Frequency")
```
```{r distribution, echo=FALSE}
ggplotly(shots.distribution)
```

##### 4. Compare actual goals and expected goals across different matches for "Manchester City WFC"
Drawn at threshold = 0.3.
```{r message=FALSE, warning=FALSE}
# Compare actual goals and expected goals across different matches for "Manchester City WFC"
# at threshold = 0.3
actual.to.xG <- data %>% filter(team.name == "Manchester City WFC") %>%
  group_by(match_id)%>%
  mutate(xG.as.goals = ifelse(xG >= 0.3, 1, 0))%>%
  summarise(actual.goals = sum(is_goal), expected.goals = sum(xG.as.goals)) %>%
  melt(id = c("match_id"))%>%
  ggplot(aes(match_id, value, color = variable)) + 
  geom_smooth()+
  labs(title= "Actual Goals vs. Expected Goals for Manchester City WFC",x ="Match ID", y = "Goals Count")
```

```{r actual, echo=FALSE}
actual.to.xG
```
 
